Transport map accelerated adaptive importance sampling, and application to inverse problems arising from multiscale stochastic reaction networks

### 2019-01-15 TETAIS.tex
Cover letter added plus minor amendments

\title{Transport map accelerated adaptive importance sampling, and application to inverse problems arising from
  multiscale stochastic reaction networks}
\begin{document}
\maketitle
\begin{abstract}
In many applications, Bayesian inverse problems can give rise to
probability distributions which contain complexities due to the
Hessian varying greatly across parameter space. This complexity often manifests itself as lower
dimensional manifolds on which the likelihood function is
invariant, or varies very little. This can be due to trying to infer
unobservable parameters, or due to sloppiness in the model which is
being used to describe the data. In such a situation, standard
sampling methods for characterising the posterior distribution, which
do not incorporate information about this structure, will be highly
inefficient. Moreover, most methods are inherently serial in nature,
and as such are not exploiting the parallelised  nature of modern
computer infrastructure. In this paper, we seek to develop an approach to
tackle this problem, using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds.

We demonstrate the approach by considering inverse problems arising
from partially observed stochastic reaction networks. In particular,
we consider systems which exhibit multiscale behaviour, but for which
only the slow variables in the system are observable. We demonstrate
that certain multiscale approximations lead to more consistent
approximations of the posterior than others.
\end{abstract}

\section{Introduction}
The importance of Markov chain Monte Carlo (MCMC) methods is becoming
increasingly apparent, in a world replete with datasets which need to be
combined with complex dynamical systems in order for us to be able
make progress in a range of scientific fields. Different MCMC methods
have been designed with different challenges in mind, for example high dimensional
targets\cite{cotter2013mcmc}, or large data
sets\cite{bierkens2016zig}. 
Other methods include piecewise deterministic Markov processes
\cite{bouchard2018bouncy},
methods which are able to make very large moves informed by the
data\cite{teh2016consistency,duane1987hybrid}, methods which are able to exploit
the geometry of the target\cite{girolami2011riemann}. Population Monte Carlo
(pMC) methods use efficient importance sampling proposals informed by an
ensemble of particles which have already learned about the structure
of the target\cite{cappe2012population,cappe2008adaptive,douc2007convergence,douc2007minimum, martino2017layered, cornuet2012adaptive, martino2015adaptive}. These methods are particularly useful in the context of
low dimensional target distributions which have complex\footnote{We
  use ``complex'' with the meaning ``complicated'' rather than
  containing imaginary numbers.} structure, for
which traditional Metropolis-Hastings methods tend to struggle, for
example when the target is multimodal or if the density is
concentrated on a lower-dimensional manifold. Moreover, they are an
excellent candidate for parallelisation, since the task of computing
the likelihood for each particle in the ensemble can be distributed
across a set of processors. This is of utmost importance in an age
where increases in computational power are most likely to come from
large-scale parallelisation as opposed to the large speed-ups in
individual processors that we saw throughout the 20th century.

In \cite{cotter2015parallel}, we explored ideas from adaptive MCMC within a
pMC-type structure, in order to construct an adaptive importance
sampler which could automatically adapt to optimise
efficiency. Alongside this, we explored the use of state-of-the-art
resamplers\cite{reich2013nonparametric}, and a greedy approximation (the
approximate multinomial resampler) in order to improve the quality of
the importance sampler proposals.

Other approaches to challenging posterior structure have been
proposed, including the work of Marzouk and Parno\cite{parno2014transport}, who
demonstrated how invertible transport maps can be constructed which
map the target distribution close to an easily-explored reference
density, such as a $d$-dimensional Gaussian. Since the map is
invertible, Gaussian proposal densities on the reference space can be
mapped to highly informed moves on the parameter space. A related idea
is the use of the anisotropic diffusion maps\cite{singer2008non}, in
which a map which describes the slow manifold is constructed using
principle component analysis. These ideas have also been applied to
stochastic reaction
networks\cite{singer2009detecting,dsilva2016data}. In
\cite{chiavazzo2017intrinsic}, these maps were used to propagate
data-mining into unexplored regions which were predicted to be close
to the lower-dimensional manifold. The idea is similar to that of the
Lamperti or Girsanov transformations for SDEs\cite{flandrin2003stationarity,gardiner2009stochastic},  the aim being to map complex
processes onto a standard one which is well understood, through a
change of measure.

@@

The partial observability of these systems lead to only
partial-observability of the unknown parameters in the system, for
example the reaction rates of the fast reactions. The resulting
multiscale approximations, and the resulting approximations of the
likelihoods arising from the inverse problem for such a network, then
often exhibit 
insensitivity to changes in certain directions of
parameter space. This causes problems when aiming to sample from such
a target, since the density is concentrated on a manifold whose
tangential vectors point in directions of insensitivity. This
\emph{sloppiness}\cite{gutenkunst2007universally,apgar2010sloppy} of the likelihood is a challenging problem
that appears in many applications in science and engineering\cite{constantine2014active}.

@@

In Section \ref{sec:ETAIS}, we will briefly reintroduce ETAIS. In Section \ref{sec:map} we show how an appropriate transport map can
be constructed from importance samples which maps the posterior close
to a reference Gaussian measure. In Section \ref{sec:TETAIS} we show
how such a map can be incorporated into a sophisticated parallel MCMC
infrastructure in order to accelerate mixing, and improve stability. In Section
\ref{sec:conv} we seek to show the advantages of this approach through
the analysis of a challenging test problem. In Section
\ref{sec:multi} we consider how likelihoods can be approximated using
 multiscale methodologies in order to carry out inference for
multiscale and/or partially observed stochastic reaction networks. In
Section \ref{sec:num} we present some numerical examples, which serve
to demonstrate the increased efficiency of the described sampling
methodologies, as well as investigating the posterior approximations
discussed in the previous section. We conclude with a discussion in
Section \ref{sec:conc}.

\section{Ensemble Transport Adaptive Importance Sampling}\label{sec:ETAIS}
Ensemble transport adaptive importance sampling (ETAIS)\cite{cotter2015parallel} is a variant of pMC\cite{cappe2012population}, which is a family of methods which are based
on importance sampling. In importance sampling, we attempt to
characterise a target density through sampling from that
density. However, the target density $\pi$ itself is often too complicated to sample
from directly, so we instead sample from a
proposal density $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
sampling from a different distribution to $\pi$. Monte Carlo estimates
using a sample of size $N$
of a function $f$ with respect to $\pi$ can then be made through the
formula
\[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
  w_kf(\theta^{(k)}),\]
where $\bar{w} = \sum_{k=1}^N w_k$ is the total weight.
This method works well when $\pi$ and $\chi$ are close, but can be
excruciatingly slow when they are not. The idea behind pMC methods is
to construct a good proposal distribution, either from the entire
history of the algorithm up to the current point, or to use the
current state of a whole ensemble of $M$ particles in the system.

In ETAIS, the proposal distribution $\chi$ is chosen to be the equally
weighted mixture of any choice of MCMC proposal kernel, evaluated at
each of the current particles in the system. If $\theta^{(k)} = [\theta_1^{(k)},
\theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
ensemble, and we wish to use an MCMC proposal density $q(\cdot ;
\cdot, \beta)$ then 
\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}, \beta).\]
Often the variance of the MCMC proposal kernels can be tuned using
their respective algorithmic parameters $\beta$. Good values for these
algorithmic parameters can be
found by optimising for the effective sample size of the importance
sample that is produced (see \cite{cotter2015parallel, russ2017parallel} for more details).

If the ensemble is large enough, and the chain has entered
probabilistic stationarity, then the current state of the ensemble is
a good rough discrete approximation of the target density, and in turn
$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
sample $\{(\hat{\theta}^{(k)},w^{(k)})\}$,  where $\theta^{(k)} =
[\theta_1^{(k)}, \ldots, \theta_M^{(k)}]^\top$ and $w^{(k)} =
[w_1^{(k)}, \ldots, w_M^{(k)}]^\top$. It can be advantageous to use stratified sampling of the
mixture in order to
ensure that the sample made is as representative as possible of the
target density, i.e.
\[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
for each $i = 1,2,\ldots,M$. Usually the kernel $q(\cdot ;
x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
is scaled with $\beta$. We now have a weighted sample, and it
would be inadvisable to use an equally weighted mixture of proposal
distributions from each of these points. Therefore, before starting
the next iteration, the importance sample
$\{(\theta^{(k)},w^{(k)})\}$ is resampled to produce an equally
weighted sample, ready for the next iteration of the algorithm. In
ETAIS, a state-of-the-art resampler is used, which uses optimal
transport methods to find the equally weighted discrete sample which
best represents the statistics of the importance
sample\cite{reich2013nonparametric}. For larger ensemble sizes $M$ this can become
expensive, in which case a greedy approximation of this algorithm, the
Multinomial Transformation (MT) can be
implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
$\theta^{(k+1)}$ and the algorithm is ready for the next
iteration. The ETAIS is summarised in Algorithm \ref{alg:ETAIS}, where
the importance samples $\{(\hat{\theta}^{(k)},w^{(k)})\}$ are stored as
the output.

\begin{table}[!h]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	Initialise $\theta^{(1)} \sim \mu_0$.\;
	\For{$k=1,\ldots,N$}{
		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_j^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:ETAIS_propose}\;
		Calculate $w^{(k)} = (w_1^{(k)}, \ldots, w_M^{(k)})^\top$, where
			\[
				w^{(k)}_i = \frac{\mu(\hat{\theta}^{(k)}_i)}{\chi^{(k)}(\hat{\theta}^{(k)}_i; \theta^{(k)}, \beta)}.
			\]
		\label{algline:ETAIS_weights}

@@

One problem with the ETAIS and other pMC methods can become apparent if
the target density has highly curved manifolds on which the density is
concentrated. In this
case, unless the proposal densities $q$ are informed by this local
structure, the mixture distribution proposal may not well approximate
$\pi$ without a very large ensemble size $M$, which can become
prohibitively expensive. Some methods have been
proposed\cite{douc2007minimum}
which use samples local to each particle to inform local covariance
structure.

In this paper, we investigate the use of transport maps to learn local
covariances across the whole of the domain, in order to stabilise
pMC-type methods, and make these methods more applicable to a wider
range of more challenging inference problems.

@@

Proposing new states in this way allows us to make large steps around complex probability distributions.
It is also feasible in this framework to assume that the reference
density is close enough to a standard Gaussian that we can efficiently
propose moves using a proposal distribution which is independent of
the current state, for example sampling from the reference Gaussian
itself, or a slightly more diffuse distribution.

In this Section we outline the methodology in
\cite{parno2014transport} for approximately coupling the target,
$\mu_{\theta}$, with the reference distribution, $\mu_r$, and
show how the map can be constructed using a weighted sample
and hence how we can incorporate the map into importance sampling schemes.

@@

\begin{dfn}[Target and Reference Space]
	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target measure $\mu_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference measure $\mu_r$.
\end{dfn}

@@

The regularisation term in Equation~\eqref{eq:blas_map_optim} can be approximated using Parseval's identity,
\[
	\frac{1}{\bar{w}}\sum\limits_{k=1}^K \! w_k
        (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)})
        \xrightarrow[K \to \infty]{}
		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\mu_\theta =
		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
\]
where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\mu_\theta$; however this
approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples.% If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.

@@

\section[Transport map MCMC]{Transport map usage in ETAIS
  and other pMC algorithms}\label{sec:TETAIS}

@@

The convergence analysis has been performed for the constrained
approximation of the posterior distribution. We now look at the
differences in the approximations of the posterior distribution
arising in this problem between the CMA and QEA. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
\[
	\hat{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{\alpha}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
\]
This difference in the denominator causes a shift in the manifold on
which the posterior is concentrated, as can be seen in
Figure~\ref{fig:chem_diff}. The figure shows the difference in
posteriors densities
\begin{equation}\label{eqn:chem_diff}
	\text{diff}(\mu^{\text{CMA}}, \mu^{\text{QEA}}) = \pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
\end{equation}

@@

One major issue with pMC-type algorithms is the curse of
dimensionality. This curse is two-fold, since as the target
distribution's dimension is increased, a larger ensemble size is often
required in order to capture the addition complexity of the density
that the higher dimension enables. Furthermore, the cost of the
resampling step within the algorithm grows with the ensemble size. For
example, the ETPF algorithm is $\mathcal{O}(n^2)$. However, the
simplification that the transport map makes has some potential to go
some way to tackling this problem.

>>>>

\title{Transport map accelerated adaptive importance sampling, and application to inverse problems arising from
  multiscale stochastic reaction networks}
\begin{document}
\maketitle
\begin{abstract}
In many applications, Bayesian inverse problems can give rise to
probability distributions which contain complexities due to the
Hessian varying greatly across parameter space. This complexity often manifests itself as lower
dimensional manifolds on which the likelihood function is
invariant, or varies very little. This can be due to trying to infer
unobservable parameters, or due to sloppiness in the model which is
being used to describe the data. In such a situation, standard
sampling methods for characterising the posterior distribution, which
do not incorporate information about this structure, will be highly
inefficient.

In this paper, we seek to develop an approach to
tackle this problem when using adaptive importance sampling methods,
by using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds. This approach is applicable to a whole range of
problems for which Monte Carlo Markoc chain (MCMC) methods mix slowly.

We demonstrate the approach by considering inverse problems arising
from partially observed stochastic reaction networks. In particular,
we consider systems which exhibit multiscale behaviour, but for which
only the slow variables in the system are observable. We demonstrate
that certain multiscale approximations lead to more consistent
approximations of the posterior than others. The use of optimal
transport maps stabilises the ensemble transform adaptive importance
sampling (ETAIS) method, and allows for efficient sampling with
smaller ensemble sizes.
\end{abstract}

\section{Introduction}
The importance of Markov chain Monte Carlo (MCMC) methods is becoming
increasingly apparent, in a world replete with datasets which need to be
combined with complex dynamical systems in order for us to be able
make progress in a range of scientific fields. Different MCMC methods
have been designed with different challenges in mind, for example high dimensional
targets\cite{cotter2013mcmc}, or large data
sets\cite{bierkens2016zig}. New types of stochastic processes
have been considered, such as piecewise deterministic Markov processes
\cite{bouchard2018bouncy}.
Other methods are able to make very large moves informed by the
data\cite{teh2016consistency,duane1987hybrid}, and others are able to exploit
the geometry of the target\cite{girolami2011riemann}. Population Monte Carlo
(pMC) methods use efficient importance sampling proposals informed by an
ensemble of particles which have already learned about the structure
of the target\cite{cappe2012population,cappe2008adaptive,douc2007convergence,douc2007minimum, martino2017layered, cornuet2012adaptive, martino2015adaptive}. These methods are particularly useful in the context of
low dimensional target distributions which have complex\footnote{We
  use ``complex'' with the meaning ``complicated'' rather than
  containing imaginary numbers.} structure, for
which traditional Metropolis-Hastings methods tend to struggle, for
example when the target is multimodal or if the density is
concentrated on a lower-dimensional manifold. Moreover, they are an
excellent candidate for parallelisation, since the task of computing
the likelihood for each particle in the ensemble can be distributed
across a set of processors. This is of utmost importance in an age
where increases in computational power are most likely to come from
large-scale parallelisation as opposed to the large speed-ups in
individual processors that we saw throughout the 20th century.

In \cite{cotter2015parallel}, we explored ideas from adaptive MCMC
within adaptive importance samplers, in order to construct a method which could automatically adapt to optimise
efficiency. Alongside this, we explored the use of state-of-the-art
resamplers\cite{reich2013nonparametric}, and the multinomial transform
(MT), a greedy approximation of the ensemble transform method, in order to improve the quality of
the importance sampler proposal distribution.

Other approaches to challenging posterior structure have been
proposed, including the work of Marzouk and Parno\cite{parno2014transport}, who
demonstrated how invertible transport maps can be constructed which
map the target distribution close to an easily-explored reference
density, such as a $d$-dimensional Gaussian. Since the map is
invertible, Gaussian proposal densities on the reference space can be
mapped to highly informed moves on the parameter space. A related idea
is the use of the anisotropic diffusion maps\cite{singer2008non}, in
which a map which describes the manifold is constructed using
principle component analysis. These ideas have also been applied to
stochastic reaction
networks\cite{singer2009detecting,dsilva2016data}. In
\cite{chiavazzo2017intrinsic}, these maps were used to propagate
data-mining into unexplored regions which were predicted to be close
to the lower-dimensional manifold. The idea is similar to that of the
Lamperti or Girsanov transformations for SDEs\cite{flandrin2003stationarity,gardiner2009stochastic},  the aim being to map complex
processes onto a standard one which is well understood, through a
change of measure.

@@

The partial observability of these systems lead to only
partial-observability of the unknown parameters in the system, for
example the reaction rates of the fast reactions. The resulting
multiscale approximations, and the resulting approximations of the
likelihoods arising from the inverse problem for such a network, 
often exhibit 
insensitivity to changes in certain directions of
parameter space. This causes problems when aiming to sample from such
a target, since the density is concentrated on a manifold whose
tangential vectors point in directions of insensitivity. This
\emph{sloppiness}\cite{gutenkunst2007universally,apgar2010sloppy} of the likelihood is a challenging problem
that appears in many applications in science and engineering\cite{constantine2014active}.

@@

In Section \ref{sec:ETAIS}, we will briefly reintroduce the ensemble
transport adaptive importance sampling (ETAIS) method. In Section \ref{sec:map} we show how an appropriate transport map can
be constructed from importance samples which maps the posterior close
to a reference Gaussian measure. In Section \ref{sec:TETAIS} we show
how such a map can be incorporated into a sophisticated parallel MCMC
infrastructure in order to accelerate mixing, and improve stability. In Section
\ref{sec:conv} we seek to show the advantages of this approach through
the analysis of a challenging test problem. In Section
\ref{sec:multi} we consider how likelihoods can be approximated using
 multiscale methodologies in order to carry out inference for
multiscale and/or partially observed stochastic reaction networks. In
Section \ref{sec:num} we present some numerical examples, which serve
to demonstrate the increased efficiency of the described sampling
methodologies, as well as investigating the posterior approximations
discussed in the previous section. We conclude with a discussion in
Section \ref{sec:conc}.

\section{Ensemble Transport Adaptive Importance Sampling}\label{sec:ETAIS}
Ensemble transport adaptive importance sampling
(ETAIS)\cite{cotter2015parallel} is an adaptive importance sampling framework
which uses an ensemble of particles and state-of-the-art resampling
methods to construct a mixture proposal distribution which closely
matches the target distribution.

In importance sampling, we aim to sample from a target $\pi$, by
sampling from a chosen proposal distribution $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
sampling from a different distribution to $\pi$. Monte Carlo estimates
using a sample of size $N$
of a function $f$ with respect to $\pi$ can then be made through the
formula
\[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
  w_kf(\theta^{(k)}),\]
where $\bar{w} = \sum_{k=1}^N w_k$ is the total weight.
This method works well when $\pi$ and $\chi$ are close, but can be
excruciatingly slow when they are not. The idea behind adaptive
importance sampling methods is
to construct a good proposal distribution, either from the entire
history of the algorithm up to the current point, or to use the
current state of a whole ensemble of $M$ particles in the system.

In ETAIS, the proposal distribution $\chi$ is chosen to be the equally
weighted mixture of set of proposal kernels. If $\theta^{(k)} = [\theta_1^{(k)},
\theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
ensemble, and we wish to use a mixture of the proposal density $q(\cdot ;
\cdot, \beta)$ then 
\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)},
\beta).\]
Usually the kernel $q(\cdot ;
x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
scales with $\beta$. Good values for the
algorithmic parameter(s) $\beta$ can be
found by optimising for the effective sample size of the importance
sample that is produced (see \cite{cotter2015parallel, russ2017parallel} for more details).

If the ensemble is large enough, and the chain has entered
probabilistic stationarity, then the current state of the ensemble is
a good rough discrete approximation of the target density, and in turn
$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
sample $\{(\hat{\theta}^{(k)},w^{(k)})\}$,  where $\theta^{(k)} =
[\theta_1^{(k)}, \ldots, \theta_M^{(k)}]^\top$ and $w^{(k)} =
[w_1^{(k)}, \ldots, w_M^{(k)}]^\top$. It can be advantageous to use stratified sampling of the
mixture in order to
ensure that the sample made is as representative as possible of the
target density, i.e.
\[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
for each $i = 1,2,\ldots,M$.  We now have a weighted sample, and it
would be inadvisable to use an equally weighted mixture of proposal
distributions from each of these points. Therefore, before starting
the next iteration, the importance sample
$\{(\theta^{(k)},w^{(k)})\}$ is resampled to produce an equally
weighted sample, ready for the next iteration of the algorithm. In
ETAIS, a state-of-the-art resampler is used, which uses optimal
transport methods to find the equally weighted discrete sample which
best represents the statistics of the importance
sample\cite{reich2013nonparametric}. For larger ensemble sizes $M$ this can become
expensive, in which case a greedy approximation of this algorithm, the
Multinomial Transformation (MT) can be
implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
$\theta^{(k+1)}$ and the algorithm is ready for the next
iteration. The ETAIS is summarised in Algorithm \ref{alg:ETAIS}, where
the importance samples $\{(\hat{\theta}^{(k)},w^{(k)})\}$ are stored as
the output.

\begin{table}[!h]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	Initialise $\theta^{(1)} \sim \pi_0$.\;
	\For{$k=1,\ldots,N$}{
		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_j^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:ETAIS_propose}\;
		Calculate $w^{(k)} = (w_1^{(k)}, \ldots, w_M^{(k)})^\top$, where
			\[
				w^{(k)}_i = \frac{\pi(\hat{\theta}^{(k)}_i)}{\chi^{(k)}(\hat{\theta}^{(k)}_i; \theta^{(k)}, \beta)}.
			\]
		\label{algline:ETAIS_weights}

@@

One problem with the ETAIS and other similar methods can become apparent if
the target density has highly curved manifolds on which the density is
concentrated. In this
case, unless the proposal densities $q$ are informed by this local
structure, the mixture distribution proposal may not well approximate
$\pi$ without a very large ensemble size $M$, which can become
prohibitively expensive. Some methods have been
proposed\cite{douc2007minimum}
which use samples local to each particle to inform local covariance
structure.

In this paper, we investigate the use of transport maps to learn local
covariances across the whole of the domain, in order to stabilise
adaptive importance sampling methods, and make these methods more applicable to a wider
range of more challenging inference problems.

@@

Proposing new states in this way allows us to make large, informed
moves, even when sampling from probability distributions with complex structure.
If a very good such map can be found, it is also possible that the
push forward of the posterior density through the map is close enough
to the reference Gaussian that we can efficiently
propose moves using a proposal distribution which is independent of
the current state, for example sampling from the reference Gaussian
itself, or a slightly more diffuse distribution.

In this Section we outline the methodology in
\cite{parno2014transport} for approximately coupling the target
measure with density
$\pi_{\theta}$, with the reference distribution with density $\pi_r$, and
show how the map can be constructed using a weighted sample
and hence how we can incorporate the map into importance sampling schemes.

@@

\begin{dfn}[Target and Reference Space]
	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target density $\pi_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference density $\pi_r$.
\end{dfn}

@@

The regularisation term in Equation~\eqref{eq:blas_map_optim} can be approximated using Parseval's identity,
\[
	\frac{1}{\bar{w}}\sum\limits_{k=1}^K \! w_k
        (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)})
        \xrightarrow[K \to \infty]{}
		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\pi_\theta =
		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
\]
where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\pi_\theta$; however this
approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples.% If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.

@@

\section[Transport map MCMC]{Transport map usage in ETAIS
  and other importance sampling algorithms}\label{sec:TETAIS}

@@

The convergence analysis has been performed for the constrained
approximation of the posterior distribution. We now look at the
differences in the approximations of the posterior distribution
arising in this problem between the CMA and QEA. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
\[
	\hat{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{\alpha}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
\]
This difference in the denominator causes a shift in the manifold on
which the posterior is concentrated, as can be seen in
Figure~\ref{fig:chem_diff}. The figure shows the difference in
posteriors densities
\begin{equation}\label{eqn:chem_diff}\pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
\end{equation}

@@

One major issue with importance sampling-based algorithms is the curse of
dimensionality. This curse is two-fold, since as the target
distribution's dimension is increased, a larger ensemble size is often
required in order to capture the addition complexity of the density
that the higher dimension enables. Furthermore, the cost of the
resampling step within the algorithm grows with the ensemble size. For
example, the ETPF algorithm is $\mathcal{O}(n^2)$. However, the
simplification that the transport map makes has some potential to go
some way to tackling this problem.
